---
title: "Lung Cancer Model"
---

Lung cancer is the leading cause of cancer deaths in the United States. While previous research has demonstrated a clear association between tobacco use and lung cancer, many new cases of lung cancer arise amongst non-smokers. We hypothesized that proximity to chemical waste release may increase the risk for lung cancer. Furthermore, proximity may increase exposure to airborne waste release.
  
We investigated the relationship between chemical release and lung cancer incidence at the county level for seven known carcinogens. Linear regression models were fitted with lung cancer incidence (per 100,000 individuals per year) as the main response and chemical release (log-pounds by year) as the main effect. We adjusted for two potential confounders, median income$^*$ as a proxy for socioeconomic status and prevalence of current or previous smokers$^{**}$. Based on our findings, we conclude that acrylonitrile, epichlorohydrin, ethyl-acrylate, formaldehyde, trans-1,3-dichloropropene, 1,3-butadiene release are significant predictors of lung cancer incidence (p-value < 0.05). In our analysis, median income and smoking prevalence were not significant confounders for epichlorohydrin and trans-1,3-dichloropropene release. 
  
Lastly, vinyl-acetate release approaches significance (p-value = 0.07) as a predictor of lung cancer incidence, adjusting for median income. In separate analyses, we find that smoking prevalence and vinyl acetate release are collinear parameters. This finding suggests that there is an underlying relationship that makes the variables non-independent in nature. Indeed, this may be the case: In a mass spectrometry [analysis](https://www.ncbi.nlm.nih.gov/pubmed/12433113) of cigarette data , it was found that vinyl acetate is a component of cigarette smoke. As such, we removed smoking prevalence as a potential confounder in our model. <br>
  
$^*$ median income (in thousands) was adjusted for at the county level by year<br>
$^{**}$ percent of current/previous smokers was adjusted for at the state level by year
  
### Regression Models for Lung Cancer Incidence
  
The table below shows the slope coefficients for each chemical in our linear models.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)

# clean data by mapping to lowercase and replacing spaces with underscore
# filtered where values of release was 0
cancer_county_chem_pop <- readRDS('./data/cancer_county_chem_pop.rds') %>% 
  janitor::clean_names() %>% 
  ungroup() %>% 
  filter(
    total_rel_summ != 0,
    pop_est > 50000) %>% 
  mutate(
    chemical = map(chemical, tolower),
    chemical = str_replace(chemical, " ", "_"),
    total_rel_log = log(total_rel_summ))

# only interested in chemicals that appear at least 10 times, for lung cancer only
# use inner join
lung_chemical = cancer_county_chem_pop %>% 
  filter(cancer == "lung") %>% 
  group_by(chemical) %>% 
  summarise(freq = n()) %>% 
  arrange(freq) %>% 
  filter(freq >= 10) %>% 
  dplyr::select(chemical)
lung_chemical2 = cancer_county_chem_pop %>% 
  filter(cancer == "lung")
  
cancer_chem_lung = dplyr::inner_join(lung_chemical, lung_chemical2, by = "chemical")

# Spread with each chemical release as separate variable, log of total release as value
lung_spread = cancer_chem_lung %>% 
  dplyr::select(-total_rel_summ) %>% 
  spread(key = chemical, value = total_rel_log) %>% 
  janitor::clean_names()

# link income data
census_income_df <- readRDS('./data/census_income_pop_tidy.rds') %>% 
  janitor::clean_names() %>% 
  separate(area_name, into = c("county", "st"), sep = ",") %>% 
  filter(st != "NA") %>% 
  map_df(tolower) %>% 
  dplyr::select(year = year_inc, st, county, med_income) %>% 
  mutate(
    year = as.integer(year),
    st = str_replace(st, " ", ""))
# join income and lung cancer data
# incidence of cancer per million
lung_spread_income = dplyr::left_join(lung_spread, census_income_df, by = c("county" = "county", "year" = "year", "st" = "st")) %>% 
  dplyr::select(med_income, everything()) %>%
  arrange(desc(med_income)) %>% 
  mutate(
    med_income = as.numeric(med_income),
    incidence_100k = round((prevalence * 100000), digits = 0)) %>% 
  dplyr::select(incidence_100k, everything())

tobacco_2010_present = read_csv(file = "./data/BRFD_tobacco_2010_present.csv") %>% 
  filter(Response == "Never") %>% 
  dplyr::select(year = "YEAR", st = "LocationAbbr", response = "Response", percent_never = "Data_Value") %>%
  janitor::clean_names() %>% 
  mutate(percent_smoke = as.numeric(100 - percent_never),
         st = tolower(st),
         year = as.numeric(year)) %>% 
  dplyr::select(year, st, percent_smoke)
  
tobacco_2010_prior = read_csv(file = "./data/BRFD_tobacco_2010_prior.csv") %>% 
  filter(Response == "Never") %>% 
  dplyr::select(year = "YEAR", st = "LocationAbbr", response = "Response", percent_never = "Data_Value") %>%
  janitor::clean_names() %>% 
  mutate(percent_smoke = as.numeric(100 - percent_never),
         st = tolower(st),
         year = as.numeric(year)) %>% 
  dplyr::select(year, st, percent_smoke)
tobacco = bind_rows(tobacco_2010_present, tobacco_2010_prior)
lung_income_tobacco = left_join(lung_spread_income, tobacco, by = c("year" = "year", "st" = "st")) %>%
  dplyr::select(percent_smoke, everything())

# Fit for formaldehyde
fit_formaldehyde = lung_income_tobacco %>% 
  lm(incidence_100k ~ formaldehyde + percent_smoke + log(med_income), data = .) 
table1 = summary(fit_formaldehyde) %>% 
  broom::tidy() %>% knitr::kable()
chem_1 = summary(fit_formaldehyde) %>% 
  broom::tidy() %>% 
  filter(term == "formaldehyde")
# Fit for acrylonitrile
fit_acrylonitrile = lung_income_tobacco %>% 
  lm(incidence_100k ~ acrylonitrile + percent_smoke + med_income, data = .) 
table2 = summary(fit_acrylonitrile) %>% 
  broom::tidy() %>% knitr::kable()
chem_2 = summary(fit_acrylonitrile) %>% 
  broom::tidy() %>%
  filter(term == "acrylonitrile")
# Fit for x1_3_butadiene
fit_butadiene = lung_income_tobacco %>% 
  lm(incidence_100k ~ x1_3_butadiene + percent_smoke + med_income, data = .) 
table3 = summary(fit_butadiene) %>% 
  broom::tidy() %>% knitr::kable()
chem_3 = summary(fit_butadiene) %>% 
  broom::tidy() %>% 
  filter(term == "x1_3_butadiene") %>% 
  mutate(term = str_replace(term, "x1_3_butadiene", "1,3-butadiene"))
# Fit for epichlorohydrin
fit_epichlorohydrin = lung_income_tobacco %>% 
  lm(incidence_100k ~ epichlorohydrin, data = .) 
table4 = summary(fit_epichlorohydrin) %>% 
  broom::tidy() %>% knitr::kable()
chem_4 = summary(fit_epichlorohydrin) %>% 
  broom::tidy() %>% 
  filter(term == "epichlorohydrin")
# Fit for vinyl_acetate
fit_vinyl_acetate = lung_income_tobacco %>% 
  lm(incidence_100k ~ vinyl_acetate + med_income, data = .) 
table5 = summary(fit_vinyl_acetate) %>% 
  broom::tidy() %>% knitr::kable()
chem_5 = summary(fit_vinyl_acetate) %>% 
  broom::tidy() %>% 
  filter(term == "vinyl_acetate") %>% 
  mutate(term = str_replace(term, "vinyl_acetate", "vinyl-acetate"))
# Fit for ethyl_acrylate
fit_ethyl_acrylate = lung_income_tobacco %>% 
  lm(incidence_100k ~ ethyl_acrylate + percent_smoke + med_income, data = .) 
table6 = summary(fit_ethyl_acrylate) %>% 
  broom::tidy() %>% knitr::kable()
chem_6 = summary(fit_ethyl_acrylate) %>% 
  broom::tidy() %>% 
  filter(term == "ethyl_acrylate") %>% 
  mutate(term = str_replace(term, "ethyl_acrylate", "ethyl-acrylate"))
# Fit for trans-1,3-dichloropropene
fit_dichloropropene = lung_income_tobacco %>% 
  lm(incidence_100k ~ trans_1_3_dichloropropene, data = .) 
table7 = summary(fit_dichloropropene) %>% 
  broom::tidy() %>% knitr::kable()
chem_7 = summary(fit_dichloropropene) %>% 
  broom::tidy() %>% 
  filter(term == "trans_1_3_dichloropropene") %>% 
  mutate(term = str_replace(term, "trans_1_3_dichloropropene", "trans-1,3-dichloropropene"))
# Beta1 Table
beta1_table = bind_rows(chem_1, chem_2, chem_3, chem_4, chem_5, chem_6, chem_7) %>% 
  arrange(term) %>% 
  mutate(
    estimate = round(estimate, digits = 2),
    std.error = round(std.error, digits = 2),
    statistic = round(statistic, digits = 2)) %>% 
  select(chemical = term, "slope estimate" = estimate, everything())
beta1_table %>% knitr::kable()
```
  
### Scatterplot
The plot below shows the relationship between lung cancer incidence and the total amount of chemicals released per county per year for the seven chemicals that we found to be significantly correlated to lung cancer incidence.
  

```{r echo=FALSE, message=FALSE, warning=FALSE}
## plotly graph
plot_ly(data = lung_spread_income, text = ~paste("State: ", st, '<br>County: ', county, '<br>Year: ', year), colors = "Set1") %>% 
  add_trace(x = ~acrylonitrile, y = ~incidence_100k, color = I("tomato"), name = 'acrylonitrile', opacity = 0.8) %>% 
  add_trace(x = ~epichlorohydrin, y = ~incidence_100k, color = I("orange"), name = 'epichlorohydrin', opacity = 0.8) %>% 
  add_trace(x = ~ethyl_acrylate, y = ~incidence_100k, color = I("seagreen"), name = 'ethyl_acrylate', opacity = 0.8) %>% 
  add_trace(x = ~formaldehyde, y = ~incidence_100k, color = I("royalblue"), name = 'formaldehyde', opacity = 0.8) %>% 
  add_trace(x = ~trans_1_3_dichloropropene, y = ~incidence_100k, color = I("mediumpurple"), name = 'trans-1,3-dichloropropene', opacity = 0.8) %>% 
  add_trace(x = ~vinyl_acetate, y = ~incidence_100k, color = I("olivedrab"), name = 'vinyl-acetate', opacity = 0.8) %>% 
  add_trace(x = ~x1_3_butadiene, y = ~incidence_100k, color = I("hotpink3"), name = '1,3-butadiene', opacity = 0.8) %>% 
  layout(title = 'Lung Cancer Incidence by Released Waste',
         xaxis = list(title = "Pounds of Annual Released Waste (log-scaled)"),
         yaxis = list(title = "Lung Cancer Incidence per 100 Thousand per Year"))
``` 
  
### How are we exposed to these chemicals?
Because we found the seven chemicals above to be correlated to cancer incidences, we wanted to figure out which industries were responsible for the most releases of each chemical. Below is a graph showing the percent contribution of different industries to the total amount released. The chemical industry is by far the largest contributor to most of the carcinogens. Our model suggests that environmental exposure to chemicals released by these industries may contribute to lung cancer incidence.

```{r echo=FALSE, message=FALSE, warning=FALSE}
tri_df = read_csv("./data/tri_df_analysis.csv")                

# filter for only lung cancer chemical's from Joy's analysis
lung_chemicals = tri_df %>%
  filter(chemical == "FORMALDEHYDE" | chemical == "ACRYLONITRILE" | chemical == "1,3-BUTADIENE" | chemical == "EPICHLOROHYDRIN" | chemical == "VINYL ACETATE" | chemical == "ETHYL ACRYLATE" | chemical =="TRANS-1,3-DICHLOROPROPENE")
# sum of total amount of chemical released per industry
industry_released = lung_chemicals %>%
  group_by(chemical, industry_sector) %>%
  summarize(total_per_industry = sum(total_releases))
# sum of total amount of chemical released
total_released = lung_chemicals %>%
  group_by(chemical) %>%
  summarize(total_release = sum(total_releases))
# chemicals relating to lung cancer with a new variable showing percent each industry contributes
lung_percent = inner_join(total_released, industry_released) %>%
  mutate(percent_released = total_per_industry/total_release) %>%
  filter(percent_released > .01) # filter to not include industries that contribute less than 1% to total released

library(viridis)
# visualization showing percent contribution from each industry
lung_percent %>%
  mutate(industry_sector = fct_reorder(industry_sector, percent_released)) %>%
  ggplot(aes(x = chemical, y = percent_released, fill = industry_sector)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_flip() +
  labs(x = "Chemical", y = "Percent Released", color = "Industry Sector", title = "Percent Contribution of Carcinogens Released by Industry")
```






